plyranges: A grammar of genomic data transformation

### 2018-09-18 paper.Rmd
avoid relational algebra before it is defined
it's Codd not Cobb

Several domain specific languages (DSLs) enable scientists to process and reason
about heterogeneous genomics data by expressing common
operations, such as range manipulation and overlap-based joins, using
the vocabulary of genomics. Their implementations either delegate
computations to a database, or operate over collections of files in
standard formats like BED. An example of the former is the Genome Query 
Language (GQL) and its distributed implementation GenAp which use a 
SQL-like syntax for fast retrieval of information of unprocessed 
sequencing data [@Kozanitis2014-va; @Kozanitis2016-bm]. Similarly, the Genometric 
Query Language (GMQL) implements a relational algebra for combining genomic datasets 
\cite{Kaitoua2017-pw}.  The command line application BEDtools develops an extensive 
algebra for performing arithmetic between two or more sets of genomic regions 
\cite{Quinlan2010-gc}. All of the aforementioned DSLs are designed to be evaluated
either at the command line or embedded in scripts for batch processing. They 
exist in a sparse ecosystem, mostly consisting of UNIX and database tools that
lack biological semantics and operate at the level of files and
database tables.

@@

For merging based on the hits, we have three modes: _inner_,
_intersect_ and _left_. The _inner_ overlap join is similar to the
conventional inner join in that there is a row in the result for every
match. A major difference is that the matching is not by identity, so
we have to choose one of the ranges from each pair. We always choose
the left range. The _intersect_ join uses the intersection instead of
the left range. Finally, the overlap _left_ join is akin to left outer
join in Cobb's relational algebra: it performs an overlap inner join
but also returns all query ranges that are not hit by the subject.

>>>>

Several domain specific languages (DSLs) enable scientists to process and reason
about heterogeneous genomics data by expressing common
operations, such as range manipulation and overlap-based joins, using
the vocabulary of genomics. Their implementations either delegate
computations to a database, or operate over collections of files in
standard formats like BED. An example of the former is the Genome Query 
Language (GQL) and its distributed implementation GenAp which use a 
SQL-like syntax for fast retrieval of information of unprocessed 
sequencing data [@Kozanitis2014-va; @Kozanitis2016-bm]. Similarly, the Genometric 
Query Language (GMQL) implements a DSL for combining genomic datasets 
\cite{Kaitoua2017-pw}.  The command line application BEDtools develops an extensive 
algebra for performing arithmetic between two or more sets of genomic regions 
\cite{Quinlan2010-gc}. All of the aforementioned DSLs are designed to be evaluated
either at the command line or embedded in scripts for batch processing. They 
exist in a sparse ecosystem, mostly consisting of UNIX and database tools that
lack biological semantics and operate at the level of files and
database tables.

@@

For merging based on the hits, we have three modes: _inner_,
_intersect_ and _left_. The _inner_ overlap join is similar to the
conventional inner join in that there is a row in the result for every
match. A major difference is that the matching is not by identity, so
we have to choose one of the ranges from each pair. We always choose
the left range. The _intersect_ join uses the intersection instead of
the left range. Finally, the overlap _left_ join is akin to left outer
join in Codd's relational algebra: it performs an overlap inner join
but also returns all query ranges that are not hit by the subject.
